---
title: "Reading and plotting deconvolution results"
author: "Chananchida Sang-aram"
date: "2026-06-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

The `runRCTD.R` script returns two files
* `*_proportions.tsv` contains the bin-by-cell type proportion information. 
* `*_doublet_info.tsv` contains information on the quality of the predictions, i.e., "singlet", "doublet_certain", "doublet_uncertain", and "reject".

Let's read in both of them.

```{r}
doublet_info <- read.table("tutorial/rctd_doublet_info.tsv",
                          header = TRUE)
head(doublet_info)

proportions <- read.table("tutorial/rctd_proportions.tsv",
                          row.names = 1, header = TRUE)
head(proportions)
```

Let's load in the original VisiumHD object to visualize the proportions spatially as we did in the paper. Each bin is represented as a barplot of proportions. We provide the plotting function in `utils/scatterbarplot_function.R` along with the explanation.

```{r, fig.width=12}
source("utils/scatterbarplot_function.R")

visium_obj_roi <- readRDS("data/Visium_HD_MouseBrain/Visium_HD_MouseBrain_016um_6951x289.rds")
square_size <- visium_obj_roi@images[["slice1.016um"]]@scale.factors$spot

# Clean up colnames
colnames(proportions) <- gsub("X[0-9]+", "", colnames(proportions))

# Color palette from ABA
color_palette <- c("IT.ETGlut" = "#FA0087",
                   "NP.CT.L6bGlut" = "#61e2a4",
                   "CTX.CGEGABA" = "#CCFF33",
                   "CTX.MGEGABA" = "#f954ee",
                   "Astro.Epen" = "#594a26",
                   "Vascular" =  "#858881",
                   "Immune" = "#825f45" )

plot_scatterbar(deconv_props = proportions,
                square_size = square_size,
                visium_obj_roi = visium_obj_roi,
                color_palette = color_palette) +
  theme(legend.text = element_text(size=12),
        legend.key.size = unit(0.5, "cm"))
```