---
title: "Reading and plotting deconvolution results"
author: "Chananchida Sang-aram"
date: "2026-06-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(Seurat)
source("utils/scatterbarplot_function.R")
```

The `runRCTD.R` script returns two files:
* `*_proportions.tsv` contains the bin-by-cell type proportions 
* `*_doublet_info.tsv` contains information about the confidence of the predictions, i.e., "singlet", "doublet_certain", "doublet_uncertain", and "reject".

Let's read in both of them.

```{r}
doublet_info <- read.table("tutorial/example_data/rctd_doublet_info.tsv",
                          header = TRUE)
head(doublet_info)

proportions <- read.table("tutorial/example_data/rctd_proportions.tsv",
                          row.names = 1, header = TRUE)
head(proportions)
```

Let's load the original Visium HD object and visualize the deconvolution results spatially, as shown in our paper. In this visualization, each bin is represented by a bar plot showing the predicted cell type proportions. The plotting function is provided in `utils/scatterbarplot_function.R`, where all function parameters are documented.

```{r, fig.width=12}
visium_obj_roi <- readRDS("tutorial/example_data/Visium_HD_MouseBrain_016um_6951x289.rds")
square_size <- visium_obj_roi@images[["slice1.016um"]]@scale.factors$spot

# Clean up colnames
colnames(proportions) <- gsub("X[0-9]+", "", colnames(proportions))

# Color palette from ABA
color_palette <- c("IT.ETGlut" = "#FA0087",
                   "NP.CT.L6bGlut" = "#61e2a4",
                   "CTX.CGEGABA" = "#CCFF33",
                   "CTX.MGEGABA" = "#f954ee",
                   "Astro.Epen" = "#594a26",
                   "Vascular" =  "#858881",
                   "Immune" = "#825f45" )

plot_scatterbar(deconv_props = proportions,
                square_size = square_size,
                visium_obj_roi = visium_obj_roi,
                color_palette = color_palette) +
  theme(legend.text = element_text(size=12),
        legend.key.size = unit(0.5, "cm"))
```


### Extra: using the scatterbarplot function on the full dataset

The example dataset covers only a small region of the tissue. When working with a complete Visium HD dataset, it is usually more visually appealing to subset the data before visualization. As an example, we will use the mouse brain FFPE dataset at 8µm resolution. The Seurat object and deconvolved proportions for the are available on [Zenodo](https://zenodo.org/records/20716044) under `full_data.zip`.

```{r}
visium_obj <- readRDS("tutorial/full_data/Visium_HD_MouseBrain_008um.rds")
visium_obj

brain_props <-  read.table("tutorial/full_data/mousebrain_ffpe_8um_proportions.tsv",
                          row.names = 1, header = TRUE)
head(brain_props)
```

We can subset the Visium HD to a region of interest (ROI) by specifying ranges of x and y coordinates. Note, however, that in the output of `Seurat::GetTissueCoordinates()`, the x coordinate corresponds to the image row index and therefore maps to the y-axis in standard Cartesian coordinates. In addition, In addition, coordinate values increase from top to bottom, rather than from bottom to top as in Cartesian plots.

```{r}
roi <- c(row_start = 6000,
         row_end = 7000,
         col_start = 4000,
         col_end = 5000)

# Visualize cells in ROI
cells_roi <- GetTissueCoordinates(visium_obj) %>%
  filter(x > roi["row_start"] & x < roi["row_end"] & 
         y > roi["col_start"] & y < roi["col_end"]) %>%
  pull(cell)

SpatialDimPlot(visium_obj,
               cells.highlight = cells_roi)

# Subset data
visium_obj_subset <- visium_obj[,Cells(visium_obj) %in% cells_roi]
```

Once the ROI has been selected, the scatterbarplot can be generated as before. The function will automatically subset the proportions table to include only bins present in the ROI Visium HD object.

```{r}
square_size <- visium_obj@images[["slice1.008um"]]@scale.factors$spot

celltype_colors <- read.csv("tutorial/full_data/celltype_colors.csv")
celltype_colors <- celltype_colors$class_color %>% setNames(celltype_colors$class)
colnames(brain_props) <- gsub("^X|\\.", "", colnames(brain_props))

plot_scatterbar(deconv_props = brain_props,
                square_size = square_size,
                visium_obj_roi = visium_obj_subset,
                color_palette = celltype_colors)
```
